---
title: "STATS 380, Assignment 2"
author: "Summer Zhang  766906916"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Question 1

```{r q1-analysis}
# (a) Load the data in R and store it in a global symbol called data.
data <- read.csv("coal.csv")

# (b) Write a function called log_likelihood() taking a single argument params for log(lambda).
log_likelihood <- function(params) {
  lambda <- exp(params[1]) 
  sum(dpois(data$deaths, lambda, log = TRUE)) 
}

# (c) Run optim() to find the value of lambda that maximises the likelihood.
opt_result <- optim(par = 0, 
                    fn = log_likelihood, 
                    method = "BFGS", 
                    control = list(fnscale = -1))

lambda_est <- exp(opt_result$par)
empirical_mean <- mean(data$deaths)

cat("Estimated lambda: ", lambda_est, "\n")
cat("Mean of deaths:   ", empirical_mean, "\n")
```

**(d) Global symbols and literals**

* **Global Symbols:** `data`, `dpois`, `sum`, `exp`. These assume that the data frame is loaded in the global environment and contains a column named deaths, and that standard base R math functions have not been overwritten.
* **Literals:** `1`, `TRUE`.

```{r q1-refactor-plot}
# (e) Refactor the code so that the data is no longer a global variable.
log_likelihood_refactored <- function(params, the_data) {
  lambda <- exp(params[1])
  sum(dpois(the_data$deaths, lambda, log = TRUE))
}

opt_result_refactored <- optim(par = 0, 
                               fn = log_likelihood_refactored, 
                               the_data = data, 
                               method = "BFGS", 
                               control = list(fnscale = -1))

# (f) Plot the data and add a red horizontal line for the estimated value of lambda.
plot(data$year, data$deaths, type = "h", 
     xlab = "year", ylab = "deaths", 
     main = "Coal Mining Deaths (1851-1962)")
abline(h = exp(opt_result_refactored$par), col = "red", lty = 2)
```

## Question 2

```{r q2-combination}
# (a) & (b) Write a new log_likelihood() function for the change-point model with error checks.
log_likelihood <- function(params, the_data) {
  if (!is.numeric(params)) stop("Argument 'params' must be numeric.")
  if (length(params) != 3) stop("Argument 'params' must be of length 3.")
  if (!is.data.frame(the_data)) stop("Argument 'the_data' must be a data frame.")
  if (!all(c("year", "deaths") %in% colnames(the_data))) {
    stop("Argument 'the_data' must have columns 'year' and 'deaths'.")
  }
  
  lambda1 <- params[1]
  lambda2 <- params[2]
  tc <- params[3]
  
  lambdas <- ifelse(the_data$year < tc, lambda1, lambda2)
  sum(dpois(the_data$deaths, lambdas, log = TRUE))
}

# (c) Write tests using stopifnot() or assertError().
library(tools)

stopifnot(is.numeric(log_likelihood(c(5, 3, 1900), data)))
assertError(log_likelihood(c(5, 3), data))
assertError(log_likelihood(letters, data))
assertError(log_likelihood(c(5, 3, 1900), 12345))
assertError(log_likelihood(c(5, 3, 1900), data.frame(x=3, y=4)))

# (d) Write a function called guess().
guess <- function() {
  c(runif(2, min = 0, max = 5), runif(1, min = 1851, max = 1962))
}

# (e) Write a function called guess_and_check().
guess_and_check <- function(the_data) {
  p <- guess()
  l <- log_likelihood(p, the_data)
  list(params = p, logl = l)
}

# (f) Using lapply() instead of a loop, call guess_and_check() 10,000 times.
guesses <- lapply(1:10000, function(x) guess_and_check(data))

# (g) Extract all of the logl values into a vector and compute the maximum.
logls <- sapply(guesses, function(x) x$logl)
top <- max(logls)
```

## Question 3

**(a)**
Commit message: "Add initial script with maximum likelihood and change-point model functions."

```{r q3-simulation, results='hide'}
# (b) & (c) Implement the acceptance procedure using a while loop and print progress.
t_c <- numeric(0)

while (length(t_c) < 1000) {
  k <- sample(1:10000, 1)
  prob <- exp(logls[k] - top)
  
  if (runif(1) < prob) {
    t_c <- c(t_c, guesses[[k]]$params[3])
    cat("Accumulated good guesses:", length(t_c), "\r")
  }
}
cat("\n")

# (d) Plot a histogram of t_c with 100 breaks.
hist(t_c, breaks = 100, 
     main = "Plausible t_c values", 
     xlab = "t_c", 
     col = "lightgray")
```

**(e) Git commit history**

```text
* 8f3c1a4 Plot histogram for accepted t_c guesses
* 5e2d8b1 Add acceptance sampling loop with console progress tracking
* 1b9e4f2 Implement random generation and validation for change-point model
* 9a3c1f0 Add initial script with maximum likelihood and change-point model functions
```